---
title: "Lab 10: Regular Expressions"
subtitle: "Stat 133, Fall 2019"
author: "Gaston Sanchez"
output: github_document
---

> ### Learning Objectives
>
> - String manipulation
> - Base R functions for strings
> - Work with the package `"stringr"`
> - String manipulation
> - More regular expressions
> - A bit of data cleaning


### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab10-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab10-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.
- Due date displayed in the syllabus (see github repo).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

------

# Part 1) Basics of String Manipulation*

In this second part of the lab,  you will be using the row names of the data frame `USArrests` (this data comes already in R):

```{r usarrests}
head(USArrests)

states <- rownames(USArrests)
head(states)
```

Here are some functions that you may need to use in this lab:

- `nchar()`
- `tolower()`
- `toupper()`
- `casefold()`
- `paste()`
- `paste0()`
- `substr()`


### Number of characters

`nchar()` allows you to count the number of characters in a string. Use it 
on `states` to get the number of characters of each state:


### Case folding

There are 3 functions to do case-folding: `tolower()`, `toupper()`, and `casefold()`. Apply each function on `states` to see what happens


### Length of State Names

We just saw how to use `nchar()` to count the number of characters in each 
state name:

```{r num-chars}
# number of charcaters
num_chars <- nchar(states)
```

Use the vector `num_chars` to obtain a frequency table called `char_freqs`, 
and then plot the frequencies with a bar chart:


### Pasting strings

R provides the `paste()` function. This function allows you to __paste__ 
(i.e. append, concatenate) character vectors separated by a blank space:

```{r pumpkin}
paste('Pumpkin', 'Pie')
```

You can give it any number of vector inputs

```{r abcde}
paste('a', 'b', 'c', 'd', 'e')
```

You can change the separator with `sep`

```{r abcde-sep}
paste('a', 'b', 'c', 'd', 'e', sep = '-')
```

`paste()` is vectorized:

```{r vectorized}
paste('a', 1:5, sep = '.')
```

There's a special wrapper around `paste()` called `paste0()` which is 
equivalent to `paste(..., sep = "")`

```{r paste0}
# paste0() -vs- paste(..., sep = "")
paste0('Pumpkin', 'Pie')
paste('Pumpkin', 'Pie', sep = '')

# paste0() is also vectorized
paste0('a', 1:5)
```


### Your Turn*: `paste()`

Use `paste()` to form a new vector with the first five states 
and their number of characters like this:

`"Alabama = 7"  "Alaska = 6"  "Arizona = 7"  "Arkansas = 8"  "California = 10"`


Now use `paste()`'s argument `collapse = ''` to _collapse_ the first five 
states like this:

`"AlabamaAlaskaArizonaArkansasCalifornia"`


### Your Turn*: Permutations

Find the number of permutations that can be formed by using all the given 
letters in `ALABAMA`. For instance: `"ALAMABA"`, `"AMALABA"`, `"AAAABML"`.



### Substrings

R provieds the function `substr()` to extract substrings in a character vector:

```{r substr-berkeley}
# extract first 3 characters
substr('Berkeley', 1, 3)
```

Use `substr()` to shorten the state names using the first 3-letters:


Use `substr()` to shorten the state names using the last 3-letters:


How would you shorten the state names using the first letter and the last
3-letters? For instance:  `"Aama"` `"Aska"` `"Aona"` `"Asas"` etc.



### Challenge

We already obtained a frequency table `char_freqs` with the counts of state 
names by number of characters. You can use those frequencies to get those 
state-names with 4-characters or 10-characters:

```{r}
# 4-char states
states[num_chars == 4]
```

```{r}
# 10-char states
states[num_chars == 10]
```

You can use `paste()` to join the 4-character states in one single string 
(i.e. _collapsing_) like this---separated by a comma and space---:

```{r}
# collapse 4-char states
paste(states[num_chars == 4], collapse = ", ")
```

Write code (using a for-loop) to obtain a 
list `states_list` containing the collapsed names by number of characters. 
If the number of characters is an even number, then the state names should be 
in capital letters. Otherwise, they should be in lower case letters.

Each list element of `states_list` must be named with the number of characters, 
followed by a dash, followed by the word `chars`: e.g. `'4-chars'`, `'5-chars'`,
etc. In total, `states_list` should have the same length as `char_freqs`.

Here's what `states_list` should look like for the first three elements:

```
$`4-chars`
[1] "IOWA, OHIO, UTAH"

$`5-chars`
[1] "idaho, maine, texas"

$`6-chars`
[1] "ALASKA, HAWAII, KANSAS, NEVADA, OREGON"
```


------

# Part 2) Intro to Regular Expressions

In this second part of the lab, we get our first contact with regular expressions (_aka_ regex).


## Data "Emotion in Text"

You'll be working with the data file 
`text-emotion.csv` available in the course github repository. 

https://raw.githubusercontent.com/ucb-stat133/stat133-labs/master/data/text-emotion.csv

The original source is the data set "Emotion in Text" from the website
Crowd Flower Data for Everyone
[https://www.crowdflower.com/data-for-everyone/](https://www.crowdflower.com/data-for-everyone/)

The file contains four columns:

- `tweet_id`: tweet identifier
- `sentiment`: class or sentiment label
- `author`: username author of the tweet
- `content`: content of the tweet

__In your `Rmd` file write R code to do computations in order to answer each of the following questions__

```{r echo = FALSE}
tweets <- read.csv('../data/text-emotion.csv', stringsAsFactors = FALSE)
```

Also, while you work on this lab you may want to look at the cheat sheets for:

- [stringr](https://github.com/ucb-stat133/stat133-cheatsheets/blob/master/stringr-cheatsheet.pdf)
- [regex](https://github.com/ucb-stat133/stat133-cheatsheets/blob/master/regular-expressions-cheatsheet.pdf)


-----

## 1) Number of characters per tweet*

- Count the number of characters in the tweet contents; create a vector for this purpose. It is possible that you find tweets containing more than 140 characters. This
has to do with the so-called _predefined XML entities_ such as 

    - `&amp;` which represents an ampersand `&`
    - `&quot;` which represent quotes `"`
    - `&lt;` which represents less-than symbol `<`
    - `&gt;` which represents greater-than symbol `>`

- Display the `summary()` of the vector obtained above.


- Likewise, graph a histogram of these counts. To plot the histogram, 
use a bin width of 5 units: 1-5, 6-10, 11-15, 16-20, etc.
In other words: the first bin involves tweets between 1 and 5 characters 
(inclusive), the second bin involves tweets containing between 6 and 10 
characters (inclusive), and so on.


- Are there any tweets with 0 characters? (_write a command that answers this question_).

- Are there any tweets with 1 character? If yes (_write commands that answer these questions_):
    + how many?
    + what is their content?
    + what is their location (i.e. index or position)?


- What is the tweet with the most characters (i.e. max length)? (_write a command that answers these questions_).
    + the number of characters
    + display its content
    + what is its location (i.e. index or position)?


-----


## 2) Author (usernames)*

According to Twitter, usernames: 

- cannot be longer than 15 characters
- can only contain alphanumeric characters (letters A-Z, numbers 0-9) with the exception of underscores (i.e. cannot contain any symbols, dashes or spaces, except underscores)
- _If you want to know more about twitter usernames, visit:_

https://help.twitter.com/en/managing-your-account/twitter-username-rules


Confirm that the values in column `author` follow each of the rules for valid usernames:

- No longer than 15 characters _(if you find usernames longer than 15 characters, display them)_

- Contain alphanumeric characters and underscores. If you find usernames containing other symbols, display them. _Hint_: The non-word character class is `"\\W"` is your friend.

- What is the number of characters of the shortest usernames? And what are the names of these authors? _(write commands to answer these questions)_


-----

## 3) Various Symbols and Strings*

- How many tweets contain at least one caret symbol `"^"` _(write a command to answer this question)_.

- How many tweets contain three or more consecutive dollar symbols `"$"` _(write a command to answer this question)_.

- How many tweets do NOT contain the characters `"a"` or `"A"` _(write a command to answer this question)_.

- Display the first 10 elements of the tweets that do NOT contain the characters `"a"` or `"A"` _(write a command to answer this question)_.

- Number of exclamation symbols `"!"`: compute a vector with the number of exclamation symbols in each tweet, and display its `summary()`.

- What's the tweet (content) with the largest number of exclamation symbols `!`? Display its content. _(write a command to answer this question)_


-----


## 4) Sentiment*

- What are the different types of sentiments (i.e. categories)? (_write a command that answers this question_)

- Compute the frequencies (i.e. counts) of each sentiment (and display these frequencies).

- Graph the relative frequencies (i.e. proportions) with a horizontal barplot (bars horizontally oriented) in decreasing order, including names of sentiment types.

- Sentiment and length of tweets: compute a table with the average length of characters per sentiment (i.e. average number of characters for `neutral` tweets, for `happy` tweets, etc.). Display this table.


-----


## 5) Optional Challenge: Hashtags

People use the hashtag symbol `#` before a relevant keyword 
or phrase in their tweet to categorize those Tweets and help them show more 
easily in Twitter search. According to Twitter, hashtags cannot contain spaces 
or punctuation symbols, and cannot start with or use only numbers.

- Count the number of (valid) hashtags in the tweet contents.

- Display such frequencies, and make a barplot of these counts
(i.e. number of tweets with 0 hashtags, with 1 hashtag, with 2 hashtags, etc).

- What is the average length of the hashtags?

- What is the most common length (i.e. the mode) of the hashtags?

```{r echo = FALSE, eval = FALSE}
# fix
hashtags <- unlist(twt_hashtags)
hashtags <- unlist(
  lapply(twt_hashtags, function(y) if (length(y)==0) NULL else y[[1]])
)

hash_length <- str_count(hashtags) - 1
summary(hash_length)
table(hash_length)
```


-----

## 6) Optional Challenge: OMG

- How many tweets contain the _individual_ strings `"omg"` or `"OMG"` _(write a command to answer this question)_. For example:

    + `omg I just saw them again`  (this would be a match)
    + `OMG I just saw them again`  (this would be a match)
    + `I just saw them again omg`  (this would be a match)
    + `I just saw them again OMG`  (this would be a match)
    + `I just saw them omg can't believe it`  (this would be a match)
    + `I just saw them OMG can't believe it`  (this would be a match)
    + `omg: I just saw them again`  (this would NOT be a match)
    + `OMG,I just saw them again`  (this would NOT be a match)
    + `I just saw them again omg!!!`  (this would NOT be a match)
    + `I just saw them again omgomgomg`  (this would NOT be a match)
    + `I just saw them again lol-omg!!!`  (this would NOT be a match)

